Possible Inconsistency in Maximum Likelihood Calculation of Ancestral States

### 2017-03-29 joint_inf.tex
reworking some bounds/notation

Our approach will be to seek bounds for the terms inside the maximization in \eqref{eq:log_likelihood_simplified} as
\begin{equation}
\label{eq:partial-likelihood}
\ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) = h_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t).
\end{equation}
We obtain an upper bound for the joint maximum of \eqref{eq:partial-likelihood}
$$
\max_{\{t,\fullAncestralStateCategories\}} \ \ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) \le
    h_{\tau^*,t^*}(\tau^*,t^*)
    + \ell^p_{\tau^*,t^*}(\hat{\fullAncestralStateCategories},\tau,\hat{t})
$$
using Gibbs's inequality
$$
h_{\tau^*,t^*}(\tau,t) \le h_{\tau^*,t^*}(\tau^*,t^*)
$$
and
$$
\{\hat{\fullAncestralStateCategories},\hat{t}\} = \arg\max_{\{\fullAncestralStateCategories,t\}} \ \ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t).
$$
Similarly, we have a lower bound as
$$
\max_{\{t,\fullAncestralStateCategories\}} \ \ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) \ge
    h_{\tau^*,t^*}(\tau,\hat{t})
    + \ell^p_{\tau^*,t^*}(\hat{\fullAncestralStateCategories},\tau,\hat{t}).
$$

@@

\begin{table}
\centering
\begin{tabular}{|l|l|l|}
\multicolumn{3}{c}{Classical}\\
    \hline
$\siteSplit_j$   &$P(\alignmentColumn=\siteSplit_j|\tau_1,t^*)$&$P(\alignmentColumn=\siteSplit_j|\tau_2,t^*)$\\
    \hline
    $\emptyset$&$1+x^2+y^2+4xy^2+x^2y^2$&$1+2xy+2xy^2+x^2y+y^3+x^2y^2$\\
    $\{1\}$    &$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
    $\{2\}$    &$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
    $\{3\}$    &$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
    $\{1,2\}$  &$1-x^2-y^2+x^2y^2$&$1+2xy-2xy^2-x^2y-y^3+x^2y^2$\\
    $\{1,3\}$  &$1+x^2+y^2-4xy^2+x^2y^2$&$1-2xy-2xy^2+x^2y+y^3+x^2y^2$\\
    $\{2,3\}$  &$1-x^2-y^2+x^2y^2$&$1-2xy+2xy^2-x^2y-y^3+x^2y^2$\\
    $\{1,2,3\}$&$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
    \hline
\multicolumn{3}{c}{Modified}\\
    \hline
$\siteSplit_j$   &$P(\alignmentColumn=\siteSplit_j|\tau_1,t)$&$P(\alignmentColumn=\siteSplit_j|\tau_2,t)$\\
    \hline
    $\emptyset$&$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
    $\{1\}$    &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
    $\{2\}$    &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
    $\{3\}$    &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
    $\{1,2\}$  &$1-x^2-y^2+x^2y^2$&$1+2xy-2xyw-x^2w-y^2w+x^2y^2$\\
    $\{1,3\}$  &$1+x^2+y^2-4xyw+x^2y^2$&$1-2xy-2xyw+x^2w+y^2w+x^2y^2$\\
    $\{2,3\}$  &$1-x^2-y^2+x^2y^2$&$1-2xy+2xyw-x^2w-y^2w+x^2y^2$\\
    $\{1,2,3\}$&$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
    \hline
\end{tabular}
\caption{Site pattern probabilities.
All values are multiplied by $1/8$.}
\label{tab:sitepatprob}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
    \multicolumn{3}{c}{$\tau=\tau_1$}\\
    \hline
    $\siteSplit_j$    & $\ancestralStateCategory_j$ & $P(\ancestralStateColumn\in\ancestralStateCategory_j|\alignmentColumn=\siteSplit_j,\tau,t)$\\
    \hline
    $\emptyset$&$\{\emptyset\}$&$(1+x)^2   (1+w)(1+y)^2$\\
     $\{1\}$    &$\{\emptyset\}$&$(1+x)(1-x)(1+w)(1+y)^2$\\
     $\{2\}$    &$\{\emptyset\}$&$(1+x)(1-x)(1+w)(1+y)^2$\\
     $\{3\}$    &$\{\emptyset\}$&$(1+x)^2   (1+w)(1+y)(1-y)$\\
    $\{1,2\}$  &$\{\emptyset,\{1,2,3\}\}$&$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
    $\{1,3\}$  &$\{\emptyset,\{2\}, \{1,2\}\}$&
    $\left\{\begin{array}{l}
                    (1+x)^2   (1+w)(1-y)^2\\
                    (1+x)^2   (1-w)(1+y)^2\\
                    (1-x)^2   (1+w)(1+y)^2
                \end{array}\right.$\\
     $\{2,3\}$  &$\{\emptyset,\{1,2,3\}\}$&$(1+x)(1-x)(1+w)(1+y)(1-y)$\\
    $\{1,2,3\}$&$\{\emptyset\}$&$(1+x)^2   (1+w)(1+y)(1-y)$\\
    \hline
    \multicolumn{3}{c}{$\tau=\tau_2$}\\
    \hline
    $\siteSplit_j$    & $\ancestralStateCategory_j$ & $P(\ancestralStateColumn\in\ancestralStateCategory_j|\alignmentColumn=\siteSplit_j,\tau,t)$\\
    \hline
 $\emptyset$       &$\{\emptyset\}$&$(1+x)^2   (1+w)(1+y)^2$\\
    $\{1\}$          &$\{\emptyset,\{1\}\}$&
    $\left\{\begin{array}{l}
                        (1+x)(1-x)(1+w)(1+y)^2\\
                        (1+x)^2   (1-w)(1+y)(1-y)
                    \end{array}\right.$\\
      $\{2\}$          &$\{\emptyset,\{1\}\}$&
    $\left\{\begin{array}{l}
                    (1+x)^2   (1+w)(1+y)(1-y)\\
                    (1+x)(1-x)(1-w)(1+y)^2
                    \end{array}\right.$\\
      $\{3\}$          &$\{\emptyset,\{2\}\}$&
    $\left\{\begin{array}{l}
                    (1+x)(1-x)(1+w)(1+y)^2\\
                    (1+x)^2   (1-w)(1+y)(1-y)
                    \end{array}\right.$\\
     $\{1,2\}$         &$\{\emptyset,\{2\},\{1,2\}\}$&
    $\left\{\begin{array}{l}
                    (1+x)(1-x)(1+w)(1+y)(1-y)\\
                    (1+x)^2   (1-w)(1+y)^2\\
                    (1+x)(1-x)(1+w)(1+y)(1-y)
                    \end{array}\right.$\\
     $\{1,3\}$         &$\{\emptyset,\{1\},\{2\},\{1,2\}\}$&
    $\left\{\begin{array}{l}
                    (1+x)^2   (1+w)(1-y)^2\\
                    (1+x)(1-x)(1-w)(1+y)(1-y)\\
                    (1+x)(1-x)(1-w)(1+y)(1-y)\\
                    (1-x)^2   (1+w)(1+y)^2
                    \end{array}\right.$\\
      $\{2,3\}$        &$\{\emptyset,\{1\},\{2\},\{1,2\}\}$&
    $\left\{\begin{array}{l}
                    (1+x)(1-x)(1+w)(1+y)(1-y)\\
                    (1-x)^2   (1-w)(1+y)^2\\
                    (1+x)^2   (1-w)(1-y)^2\\
                    (1+x)(1-x)(1+w)(1+y)(1-y)
                    \end{array}\right.$\\
     $\{1,2,3\}$       &$\{\emptyset,\{2\}\}$&
    $\left\{\begin{array}{l}
                    (1+x)^2   (1+w)(1+y)(1-y)\\
                    (1+x)(1-x)(1-w)(1+y)^2
                    \end{array}\right.$\\
    \hline
\end{tabular}
\caption{Likelihood calculations for all site patterns and ancestral state categories.
All values multiplied by $1/32$.
Likelihoods with multiple entries will have maxima determined by unknown branch length parameters.}
\label{tab:likelihoods}
\end{table}

@@

Assume $\tau^*=\tau_1$ and $t^*=\{x,y,y\}$.
To obtain an upper bound for the Farris partial likelihood---the terms in the top half of Table~\ref{tab:likelihoods}---we use the term $(1+x)^2(1+w)(1+y)^2$ for the $\{1,3\}$ split.
Since all cases have a $(1+x)^2$ term except $\{\{1\}, \{2\}, \{1,2\}, \{2,3\}\}$, and all cases except $\{\{3\}, \{1,2\}, \{2,3\}, \{1,2,3\}\}$ have a $(1+y)^2$, we can add and subtract the $\log$ of these terms weighted by each of their generating probabilities.
In other words, define
\begin{align*}
    a_{xy} &= P(\alignmentColumn=\{1\}|\tau=\tau_1,t=\{x,y,y\})
    +        P(\alignmentColumn=\{2\}|\tau=\tau_1,t=\{x,y,y\}) \\
    &\qquad + P(\alignmentColumn=\{1,2\}|\tau=\tau_1,t=\{x,y,y\})
    + P(\alignmentColumn=\{2,3\}|\tau=\tau_1,t=\{x,y,y\})
\end{align*}
and
\begin{align*}
    b_{xy} &= P(\alignmentColumn=\{3\}|\tau=\tau_1,t=\{x,y,y\}) + P(\alignmentColumn=\{1,2\}|\tau=\tau_1,t=\{x,y,y\}) \\
           &\qquad + P(\alignmentColumn=\{2,3\}|\tau=\tau_1,t=\{x,y,y\}) + P(\alignmentColumn=\{1,2,3\}|\tau=\tau_1,t=\{x,y,y\}).
\end{align*}
Then the ancestral state maximizing the likelihood is $00$ for each site pattern and the partial likelihood is bounded above
\begin{align*}
    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_1,\{x',y',w'\}) \\
    &\qquad \le (2-a_{xy})\log(1+x')+a_{xy}\log(1-x')+(2-b_{xy})\log(1+y')+b_{xy}\log(1-y')+\log(1+w') \\
    &\qquad \le (2-a_{xy})\log(2-a_{xy})+a_{xy}\log(a_{xy})+(2-b_{xy})\log(2-b_{xy})+b_{xy}\log(b_{xy})+\log(2) \\
    &\qquad \triangleq L^p_{\tau_1,\tau_1}(a_{xy}, b_{xy})
\end{align*}
where $a_{xy}$ and $b_{xy}$ are functions of the true generating branch parameters $\{x,y\}$.

@@

We construct a similar lower bound on the Felsenstein partial likelihood, though here we replace all $(1+w)$ terms with $(1-w)$.
Let
\begin{align*}
    c_{xy} &= P(\alignmentColumn=\{1\}|\tau=\tau_1,t=\{x,y,y\}) + P(\alignmentColumn=\{2\}|\tau=\tau_1,t=\{x,y,y\}) \\
           &\qquad + P(\alignmentColumn=\{3\}|\tau=\tau_1,t=\{x,y,y\}) + P(\alignmentColumn=\{1,2,3\}|\tau=\tau_1,t=\{x,y,y\})
\end{align*}
and
$$
d_{xy} = P(\alignmentColumn=\{1,3\}|\tau=\tau_1,t=\{x,y,y\}) + P(\alignmentColumn=\{2,3\}|\tau=\tau_1,t=\{x,y,y\}).
$$
The lower bounds on the likelihoods in this case take one of six forms:
Since the maxima for pairs of lower bounds above each have a symmetry in $x$ and $y$, they are bounded below by one of
\textbf{TODO}: ancestral states where these likelihoods occur

\begin{align*}
    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_2,\{x',y',w'\}) \\
    &\qquad \ge 2\log(1+x')+(2-c_{xy}-2d_{xy})\log(1+y')+(c_{xy}+2d_{xy})\log(1-y')+\log(1-w') \\
    &\qquad \ge 2\log(2)+(2-c_{xy}-2d_{xy})\log(2-c_{xy}-2d_{xy})+(c_{xy}+2d_{xy})\log(c_{xy}+2d_{xy}) \\
    &\qquad \triangleq L^{p,1}_{\tau_1,\tau_2}(c_{xy}, d_{xy}),
\end{align*}
\begin{align*}
    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_2,\{x',y',w'\}) \\
    &\qquad \ge (2-2d_{xy})\log(1+x')+2d_{xy}\log(1-x') + (2-c_{xy})\log(1+y')+c_{xy}\log(1-y')+\log(1-w') \\
    &\qquad \ge (2-2d_{xy})\log(2-2d_{xy})+2d_{xy}\log(2d_{xy}) + (2-c_{xy})\log(2-c_{xy})+c_{xy}\log(c_{xy}) \\
    &\qquad \triangleq L^{p,2}_{\tau_1,\tau_2}(c_{xy}, d_{xy})
\end{align*}
or
\begin{align*}
    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_2,\{x',y',w'\}) \\
    &\qquad \ge (2-d_{xy})\log(1+x')+d_{xy}\log(1-x') + (2-c_{xy}-d_{xy})\log(1+y')+(c_{xy}+d_{xy})\log(1-y')+\log(1-w') \\
    &\qquad \ge (2-d_{xy})\log(2-d_{xy})+d_{xy}\log(d_{xy}) + (2-c_{xy}-d_{xy})\log(2-c_{xy}-d_{xy})+(c_{xy}+d_{xy})\log(c_{xy}+d_{xy}) \\
    &\qquad \triangleq L^{p,3}_{\tau_1,\tau_2}(c_{xy}, d_{xy})
\end{align*}
where $c_{xy}$ and $d_{xy}$ are also functions of the true generating branch parameters $\{x,y\}$.

\textbf{Claim to prove}:
$$
L^{p,1}_{\tau_1,\tau_2}(c_{xy}, d_{xy}) \ge L^{p,2}_{\tau_1,\tau_2}(c_{xy}, d_{xy})
$$
and
$$
L^{p,1}_{\tau_1,\tau_2}(c_{xy}, d_{xy}) \ge L^{p,3}_{\tau_1,\tau_2}(c_{xy}, d_{xy}).
$$
I think we can use information theory results to show this.
It is definitely true---will work on it.

The values of $(x', y', w')$ maximizing $L^{p,1}_{\tau_1,\tau_2}(c_{xy}, d_{xy})$ are $(1, 1-c_{xy}-2d_{xy},0)$.

>>>>

Our approach will be to seek bounds for the terms inside the maximization in \eqref{eq:log_likelihood_simplified} as
\begin{equation}
\label{eq:partial-likelihood}
\ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) = h_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t).
\end{equation}
We obtain an upper bound for the joint maximum of \eqref{eq:partial-likelihood}
$$
\max_{t,\fullAncestralStateCategories} \ \ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) \le
    h_{\tau^*,t^*}(\tau^*,t^*)
    + \ell^p_{\tau^*,t^*}(\hat{\fullAncestralStateCategories},\tau,\hat{t})
$$
using Gibbs's inequality
$$
h_{\tau^*,t^*}(\tau,t) \le h_{\tau^*,t^*}(\tau^*,t^*)
$$
and
$$
\{\hat{t},\hat{\fullAncestralStateCategories}\} = \arg\max_{t,\fullAncestralStateCategories} \ \ell^p_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t).
$$
Similarly, we have a lower bound as
$$
\max_{t,\fullAncestralStateCategories} \ \ell_{\tau^*,t^*}(\fullAncestralStateCategories,\tau,t) \ge
    h_{\tau^*,t^*}(\tau,\hat{t})
    + \ell^p_{\tau^*,t^*}(\hat{\fullAncestralStateCategories},\tau,\hat{t}).
$$

@@

\begin{table}
\centering
\begin{tabular}{|l|l|l|}
\multicolumn{3}{c}{Classical}\\
    \hline
$\siteSplit_j$   &$P(\alignmentColumn=\siteSplit_j|\tau_1,t^*)$&$P(\alignmentColumn=\siteSplit_j|\tau_2,t^*)$\\
    \hline
    $\{\emptyset\}$&$1+x^2+y^2+4xy^2+x^2y^2$&$1+2xy+2xy^2+x^2y+y^3+x^2y^2$\\
    $\{1\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
    $\{2\}$        &$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
    $\{3\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2y+y^3-x^2y^2$\\
    $\{1,2\}$      &$1-x^2-y^2+x^2y^2$&$1+2xy-2xy^2-x^2y-y^3+x^2y^2$\\
    $\{1,3\}$      &$1+x^2+y^2-4xy^2+x^2y^2$&$1-2xy-2xy^2+x^2y+y^3+x^2y^2$\\
    $\{2,3\}$      &$1-x^2-y^2+x^2y^2$&$1-2xy+2xy^2-x^2y-y^3+x^2y^2$\\
    $\{1,2,3\}$    &$1+x^2-y^2-x^2y^2$&$1+x^2y-y^3-x^2y^2$\\
    \hline
\multicolumn{3}{c}{Modified}\\
    \hline
$\siteSplit_j$   &$P(\alignmentColumn=\siteSplit_j|\tau_1,t)$&$P(\alignmentColumn=\siteSplit_j|\tau_2,t)$\\
    \hline
    $\{\emptyset\}$&$1+x^2+y^2+4xyw+x^2y^2$&$1+2xy+2xyw+x^2w+y^2w+x^2y^2$\\
    $\{1\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
    $\{2\}$        &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
    $\{3\}$        &$1-x^2+y^2-x^2y^2$&$1-x^2w+y^2w-x^2y^2$\\
    $\{1,2\}$      &$1-x^2-y^2+x^2y^2$&$1+2xy-2xyw-x^2w-y^2w+x^2y^2$\\
    $\{1,3\}$      &$1+x^2+y^2-4xyw+x^2y^2$&$1-2xy-2xyw+x^2w+y^2w+x^2y^2$\\
    $\{2,3\}$      &$1-x^2-y^2+x^2y^2$&$1-2xy+2xyw-x^2w-y^2w+x^2y^2$\\
    $\{1,2,3\}$    &$1+x^2-y^2-x^2y^2$&$1+x^2w-y^2w-x^2y^2$\\
    \hline
\end{tabular}
\caption{Site pattern probabilities.
All values are multiplied by $1/8$.}
\label{tab:sitepatprob}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
    \multicolumn{3}{c}{$\tau=\tau_1$}\\
    \hline
    $\siteSplit_j$    & $\ancestralStateCategory_j$ & $P(\ancestralStateColumn\in\ancestralStateCategory_j|\alignmentColumn=\siteSplit_j,\tau,t)$\\
    \hline
    $\{\emptyset\}$&
    $\{\emptyset\}$&
    $(1+x)^2   (1+w)(1+y)^2$\\
     $\{1\}$    &
    $\{\emptyset\}$&
    $(1+x)(1-x)(1+w)(1+y)^2$\\
     $\{2\}$    &
    $\{\emptyset\}$&
    $(1+x)(1-x)(1+w)(1+y)^2$\\
     $\{3\}$    &
    $\{\emptyset\}$&
    $(1+x)^2   (1+w)(1+y)(1-y)$\\
    $\{1,2\}$  &
    $\{\{\emptyset\},\{1,2,3\}\}$&
    $(1+x)(1-x)(1+w)(1+y)(1-y)$\\
    $\{1,3\}$  &
    $\left\{\begin{array}{l}
                    \{\emptyset\}\\
                    \{2\}\\
                    \{1,2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)^2   (1+w)(1-y)^2\\
                    (1+x)^2   (1-w)(1+y)^2\\
                    (1-x)^2   (1+w)(1+y)^2
                \end{array}$\\
    $\{2,3\}$  &
                $\{\{\emptyset\},\{1,2,3\}\}$&
                $(1+x)(1-x)(1+w)(1+y)(1-y)$\\
    $\{1,2,3\}$&
                $\{\emptyset\}$&
                $(1+x)^2   (1+w)(1+y)(1-y)$\\
    \hline
    \multicolumn{3}{c}{$\tau=\tau_2$}\\
    \hline
    $\siteSplit_j$    & $\ancestralStateCategory_j$ & $P(\ancestralStateColumn\in\ancestralStateCategory_j|\alignmentColumn=\siteSplit_j,\tau,t)$\\
    \hline
    $\{\emptyset\}$       &$\{\emptyset\}$&$(1+x)^2   (1+w)(1+y)^2$\\
    $\{1\}$          &
    $\left\{\begin{array}{l}
                    \{\emptyset\}\\
                    \{1\}
                \end{array}\right.$&
    $\begin{array}{l}
                        (1+x)(1-x)(1+w)(1+y)^2\\
                        (1+x)^2   (1-w)(1+y)(1-y)
                    \end{array}$\\
      $\{2\}$          &
    $\left\{\begin{array}{l}
                    \{\emptyset\}\\
                    \{1\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)^2   (1+w)(1+y)(1-y)\\
                    (1+x)(1-x)(1-w)(1+y)^2
                    \end{array}$\\
      $\{3\}$          &
    $\left\{\begin{array}{l}
                    \{\emptyset\}\\
                    \{2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)(1-x)(1+w)(1+y)^2\\
                    (1+x)^2   (1-w)(1+y)(1-y)
                    \end{array}$\\
     $\{1,2\}$         &
    $\left\{\begin{array}{l}
                    \{\{\emptyset\},\{1,2\}\}\\
                    \{2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)(1-x)(1+w)(1+y)(1-y)\\
                    (1+x)^2   (1-w)(1+y)^2
                    \end{array}$\\
     $\{1,3\}$         &
    $\left\{\begin{array}{l}
                    \{\emptyset\}\\
                    \{\{1\},\{2\}\}\\
                    \{1,2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)^2   (1+w)(1-y)^2\\
                    (1+x)(1-x)(1-w)(1+y)(1-y)\\
                    (1-x)^2   (1+w)(1+y)^2
                    \end{array}$\\
      $\{2,3\}$        &
    $\left\{\begin{array}{l}
                    \{\{\emptyset\},\{1,2\}\}\\
                    \{1\}\\
                    \{2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)(1-x)(1+w)(1+y)(1-y)\\
                    (1-x)^2   (1-w)(1+y)^2\\
                    (1+x)^2   (1-w)(1-y)^2
                    \end{array}$\\
     $\{1,2,3\}$       &
    $\left\{\begin{array}{l}
                    \{\emptyset\}\\
                    \{2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x)^2   (1+w)(1+y)(1-y)\\
                    (1+x)(1-x)(1-w)(1+y)^2
                    \end{array}$\\
    \hline
\end{tabular}
\caption{Likelihood calculations for all site patterns and ancestral state categories.
All values multiplied by $1/32$.
Likelihoods with multiple entries will have maxima determined by unknown branch length parameters.}
\label{tab:likelihoods}
\end{table}

@@

Assume $\tau^*=\tau_1$ and $t^*=\{x,y,y\}$.
Define, e.g., $p_0 = P(\alignmentColumn=\{\emptyset\}|\tau=\tau_1,t=\{x,y,y\})$.
Let
$$
    a_{0} = p_{13},
$$
$$
    a_{1} = p_1+p_2+p_{12}+p_{23},
$$
$$
    a_{2} = p_3+p_{12}+p_{23}+p_{123},
$$
$$
    a_{3} = p_1+p_3+p_{12}+p_{23},
$$
$$
    a_{4} = p_2+p_{12}+p_{23}+p_{123}.
$$
Then the full likelihood is bounded above by one of three functions:
\begin{align*}
    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_1,\{x',y',w'\}) \\
    &\qquad \le (3-a_{1})\log(1+x')+(a_{1}+a_{3})\log(1-x')\\
    &\qquad \qquad +(3-a_{2})\log(1+y')+(a_{2}+a_{4})\log(1-y')\\
    &\qquad \qquad +(1-a_{0})\log(1+w')+a_{0}\log(1-w')+p_{0}\log(4)\\
    &\qquad \le (3-a_{1})\log\frac{2(3-a_{1})}{3+a_{3}}+(a_{1}+a_{3})\log\frac{2(a_{1}+a_{3})}{3+a_{3}}\\
    &\qquad \qquad +(3-a_{2})\log\frac{2(3-a_{2})}{3+a_{4}}+(a_{2}+a_{4})\log\frac{2(a_{2}+a_{4})}{3+a_{4}}\\
    &\qquad \qquad +(1-a_{0})\log(1+w')+a_{0}\log(1-w')+p_{0}\log(4)\\
    &\qquad \triangleq L^{p,1}_{\tau_1,\tau_1}(\mathbf{a}_{xy});
\end{align*}
\begin{align*}
    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_1,\{x',y',w'\}) \\
    &\qquad \le (3-a_{1}-2a_{0})\log(1+x')+(a_{1}+a_{3}+2a_{0})\log(1-x')\\
    &\qquad \qquad +(3-a_{2})\log(1+y')+(a_{2}+a_{4})\log(1-y')+\log(1+w')+p_{0}\log(4)\\
    &\qquad \le (3-a_{1}-2a_{0})\log\frac{2(3-a_{1}-2a_{0})}{3+a_{3}}+(a_{1}+a_{3}+2a_{0})\log\frac{2(a_{1}+a_{3}+2a_{0})}{3+a_{3}}\\
    &\qquad \qquad +(3-a_{2})\log\frac{2(3-a_{2})}{3+a_{4}}+(a_{2}+a_{4})\log\frac{2(a_{2}+a_{4})}{3+a_{4}}\\
    &\qquad \qquad +\log(2)+p_{0}\log(4)\\
    &\qquad \triangleq L^{p,2}_{\tau_1,\tau_1}(\mathbf{a}_{xy});
\end{align*}
\begin{align*}
    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_1,\{x',y',w'\}) \\
    &\qquad \le (3-a_{1})\log(1+x')+(a_{1}+a_{3})\log(1-x')\\
    &\qquad \qquad +(3-a_{2}-2a_{0})\log(1+y')+(a_{2}+a_{4})+2a_{0}\log(1-y')+\log(1+w')+p_{0}\log(4)\\
    &\qquad \le (3-a_{1})\log\frac{2(3-a_{1})}{3+a_{3}}+(a_{1}+a_{3})\log\frac{2(a_{1}+a_{3})}{3+a_{3}}\\
    &\qquad \qquad +(3-a_{2}-2a_{0})\log\frac{2(3-a_{2}-2a_{0})}{3+a_{4}}+(a_{2}+a_{4}+2a_{0})\log\frac{2(a_{2}+a_{4}+2a_{0})}{3+a_{4}}\\
    &\qquad \qquad +\log(2)+p_{0}\log(4)\\
    &\qquad \triangleq L^{p,3}_{\tau_1,\tau_1}(\mathbf{a}_{xy}).
\end{align*}

@@

We construct a similar lower bound on the Felsenstein partial likelihood, though here we replace all $(1+w)$ terms with $(1-w)$.
Let
$$
b_1 = p_1+p_2+p_3+p_{123}
$$
and
$$
b_2 = p_{13}+p_{23}.
$$

\textbf{TODO}: ancestral states where these likelihoods occur
Since the maxima for pairs of lower bounds above each have a symmetry in $x$ and $y$, they are bounded below by one of
\begin{align*}
    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_2,\{x',y',w'\}) \\
    &\qquad \ge 2\log(1+x')+(2-b_1-2b_2)\log(1+y')+(b_1+2b_2)\log(1-y')+\log(1-w'),
\end{align*}
\begin{align*}
    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_2,\{x',y',w'\}) \\
    &\qquad \ge (2-2b_2)\log(1+x')+2b_2\log(1-x') + (2-b_1)\log(1+y')+b_1\log(1-y')+\log(1-w')
\end{align*}
or
\begin{align*}
    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_2,\{x',y',w'\}) \\
    &\qquad \ge (2-b_2)\log(1+x')+b_2\log(1-x') + (2-b_1-b_2)\log(1+y')+(b_1+b_2)\log(1-y')+\log(1-w')
\end{align*}
where $b_1$ and $b_2$ are also functions of the true generating branch parameters $\{x,y\}$.
Taking the maximum of each yields
$$
    L^{p,1}_{\tau_1,\tau_2}(b_1, b_2) \triangleq 2\log(2)+(2-b_1-2b_2)\log(2-b_1-2b_2)+(b_1+2b_2)\log(b_1+2b_2),
$$
$$
    L^{p,2}_{\tau_1,\tau_2}(b_1, b_2) \triangleq (2-2b_2)\log(2-2b_2)+2b_2\log(2b_2) + (2-b_1)\log(2-b_1)+b_1\log(b_1)
$$
and
$$
    L^{p,3}_{\tau_1,\tau_2}(b_1, b_2) \triangleq (2-b_2)\log(2-b_2)+b_2\log(b_2) + (2-b_1-b_2)\log(2-b_1-b_2)+(b_1+b_2)\log(b_1+b_2).
$$

\textbf{Claim to prove}:
$$
L^{p,1}_{\tau_1,\tau_2}(b_1, b_2) \ge L^{p,2}_{\tau_1,\tau_2}(b_1, b_2)
$$
and
$$
L^{p,1}_{\tau_1,\tau_2}(b_1, b_2) \ge L^{p,3}_{\tau_1,\tau_2}(b_1, b_2).
$$
I think we can use information theory results to show this.
It is definitely true---will work on it.

The values of $(x', y', w')$ maximizing $L^{p,1}_{\tau_1,\tau_2}(b_1, b_2)$ are $(1, 1-b_1-2b_2,0)$.
